Graphical Tests for Power Comparison of Competing Designs

### 2012-03-29 experiment-two.tex
finished re-working the intro

% !TEX root = comparison.tex
\subsection{Comparing Distributions}
In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions. As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

>>>>

% !TEX root = comparison.tex
\subsection{Comparing Distributions}~\label{experimenttwo}
In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions. As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

### 2012-03-29 intro.tex
finished re-working the intro

Here we suggest a new approach using recent research on visual
inference \citep{buja:2009,wickham:2010}. In visual inference, real
data plots are considered to be test statistics, and these are
compared with plots of data generated from a null hypothesis using a
lineup. A lineup is similar to the police lineup, and this is where
the name comes from. The suspect (the plot of the real data) is placed
randomly in a field of plots of null data, and an impartial viewer is
asked to identify the plot that is the most different. If the viewer
identifies the suspect it lends statistical signficance to the
conclusion that the real data is not consistent with the null data.

Lineups enable specific plot design evaluation of a specific
visualization result. Different lineups are created, using the same
data, same null plots and positions within the lineup, but different
plot designs. Comparing how long and how accurately the viewers take
to report the feature of interest will assess which design is better
for the task. In addition, this approach is ideally coupled with
simulation studies, where data is constructed to match and expand on
the features of interest in the real data. For example, in a simple
situation, where the discovered feature of interest is a difference
between the centers of two distributions, simulated data would be
constructed roughly matching the distribution of the real data that
varies the distance between centers, and also spread and sample
size. Lineups of these simulated data (embedded with simulated null
data plots) are generated and evaluated along with the lineups of real
data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a guage for
the signal strength of the pattern in the real data. It provides a way
to objectively evaluate plot design.

@@

Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$.
In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots. The decoy or {\it null} plots are generated in a way that is consistent with the null hypothesis, i.e. these are plots in which we can assume that any relationship visible is there purely by chance. This gives us a reference framework against which to compare the data plot and, if we are still able to identify the data plot from the lineup because of a distinguishing feature, we have formal evidence that the feature in the data is, in fact, not just coincidental.

The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot:
this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.

We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum:
this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal.
 In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.

\section{Comparison of Designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} create data for a lineup of size $m$ by  creating  $m-1$ permutations of $y$ or, in the case of a simulation study, drawing $m-1$ samples of size $n$ (the number of rows in the data) from the null distribution. Add the original data to the lineup data randomly between 1 and $m$. }
\item{{\bf Create lineups from competing designs:} using the same data, render lineups of all competing designs. }
\item{{\bf Evaluate Lineups:} by presenting the lineups to independent observers. Assess both signal strength and time needed by individuals to come to a decision. Note that each observer should only be exposed to each lineup data once.}
\item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design. In the case that individuals were shown multiple lineups (as part of a bigger study), it is possible to correct outcome measurements for an individual's visual ability. }
\end{enumerate}

>>>>

Here we suggest a new approach using recent research on visual
inference \citep{buja:2009,wickham:2010}. In visual inference, real
data plots are considered to be test statistics, and these are
compared with plots of data generated from a null hypothesis using a
lineup. A lineup is similar to the police lineup, and this is where
the name comes from. The suspect (the plot of the real data) is placed
randomly in a field of plots of null data, and an impartial viewer is
asked to identify the plot that is the most different. If the viewer
identifies the suspect it lends statistical signficance to the
conclusion that the real data is not consistent with the null
data. Figure \ref{lineup} has an example of a lineup. Overlaid density
plots are used to display the distribution of two groups (blue, red)
of data. We are interested to know whether the real data has a
significant diffeerence between the centers of the two groups. Plot
numbered 13 contains the real data. Did you pick that one? In this
plot the blue group has a flat density whereas the density of red
group is more peaked and concentrated to the left of the blue.

\begin{figure}[htbp]
   \centering
   \includegraphics[height=0.95\linewidth]{density_1_45_15_20_13.png} 
   \caption{Example lineup using density plots: one plot of real data
     embedded with nineteen plots of null data. Which plot is
     most different from the others? (See the text for the answer.)}
   \label{lineup}
\end{figure}

The previous work \citet{buja:2009,wickham:2010} describes the use of
lineups for evaluating discoveries made in exploratory analysis. Here
we describe the use of lineups for evalating plot design for
communicating findings visually. Different lineups are created, using
the same data, same null plots and positions within the lineup, but
different plot designs. Comparing how long and how accurately the
viewers take to report the feature of interest will assess which
design is better for the task. In addition, this approach is ideally
coupled with simulation studies, where data is constructed to match
and expand on the features of interest in the real data. For example,
in a simple situation, where the discovered feature of interest is a
difference between the centers of two distributions, simulated data
would be constructed roughly matching the distribution of the real
data that varies the distance between centers, and also spread and
sample size. Lineups of these simulated data (embedded with simulated
null data plots) are generated and evaluated along with the lineups of
real data. The simulation study provides a backdrop for the real data,
which enables the broad applicability to be studied and a guage for
the signal strength of the pattern in the real data. It provides a way
to objectively evaluate plot design.

@@

%Lineup plots have been shown as an effective tool \cite{wickham:2010} for enabling us to quantitively assess the strength of the relationship between $X$ and $Y$. In a lineup plot the plot of the actual data is -- as in a police line-up -- inserted among decoy plots. The decoy or {\it null} plots are generated in a way that is consistent with the null hypothesis, i.e. these are plots in which we can assume that any relationship visible is there purely by chance. This gives us a reference framework against which to compare the data plot and, if we are still able to identify the data plot from the lineup because of a distinguishing feature, we have formal evidence that the feature in the data is, in fact, not just coincidental.

Lineups provide a way to quantify the significance of the finding
\citet{majumder:2011}. The significance depends on the size ($m$) of
the lineup ($m=20$ in Figure \ref{lineup}) and the number of
independent observers ($n$). Let $x$ be the number of observers who
correctly identify the plot of the real data. The probability that one
observer correctly identifies the real data plot, in a size $m$
lineup, by chance is $1/m$. To incorporate the results from multiple
observers think about the lineup as $m-1$ head-to-head competitions of
the data plot against $m-1$ null plots, which the data plot has to
win, is picked as the plot with the most distinct features. Assume
that the data plot is the one with the strongest signal, then the
probability that the data plot is picked by $x$ observers follows a
Beta distribution for the maximum \citet{****}:
\[
\text{signal strength} =  \left( \frac{x}{n}\right)^{1/(m-1)}.
\]
We call it signal strength, and it is analgous to power in classical hypothesis testing. Signal strength will range between 0 and 1, with higher values indicating stronger signal, more distinctive feature in the data.

*** The above paragraph needs more explanation, and a reference for the beta distribution. I don't think the Majumder paper really is the source for this result is it? It leads up to it, but this is the new work that is not submitted yet.

*** Also need to refer to Niladri's ``Where's Waldo paper'' here too.

%The strength of this evidence is determined by both the number of independent evaluations as well as the size of the lineup. Assume a lineup of size $m$ is evaluated by $n$ independent observers, out of which $x$ identify the data plot correctly, which we can use as an estimate for the probability to correctly identify the data plot: this leads us to an estimate for the probability of correctly identifying the plot of $x/n$, with 95\% confidence intervals  for $x/n$ given as $x/n \pm 1.96 \cdot \sqrt{x/n \cdot (1-x/n)}$ limited to the interval [0,1]. Note, that the resulting interval is often too large to be useful.

%We can think of a lineup as $m-1$ head-to-head comparisons of the dataplot and $m-1$ nullplots, which the dataplot all has to `win', in order to be picked as the plot with the most distinct features. Assuming that observers identify the plot with the strongest signal, the probability that data plot is picked follows a Beta distribution for the maximum: this allows us to introduce the {\it signal strength} \cite{majumder:2011} of a lineup as:

%Signal strength is limited to a range of values between 0 and 1. Higher values indicate stronger signal. In the case of a parametric situation (i.e. the feature shown in the chart can be described in parametric form as a real-valued vector $\theta$ in the data), it can be shown \cite{majumder:2011} that the signal strength is strongly correlated to the $p$-value of the corresponding statistical test for $\theta$.

Signal strength is used in our new work to evaluate the effectiveness of different plot designs. This paper describes two examples where we had real data analysis problems and decisions to make in order to communicate the results. Section \ref{designs} describes the process for comparing designs. Sections \ref{experimentone} and \ref{experimenttwo} describe the two data analysis problems and the experiments conducted to evaluate the plots designs. Section \ref{results} presents the findings of the experiments.

\section{Comparison of Designs}\label{designs}
We are going to make use of the signal strength gained from multiple viewings of a lineup in order to evaluate competing designs as follows:
\begin{enumerate}
\item{{\bf Create Lineup Data:} create data for a lineup of size $m$ by  creating  $m-1$ permutations of $y$ or, in the case of a simulation study, drawing $m-1$ samples of size $n$ (the number of rows in the data) from the null distribution. Add the original data to the lineup data randomly between 1 and $m$. }
\item{{\bf Create lineups from competing designs:} using the same data, render lineups of all competing designs. }
\item{{\bf Evaluate Lineups:} by presenting the lineups to independent observers. Assess both signal strength and time needed by individuals to come to a decision. Note that each observer should only be exposed to each lineup data once.}
\item{{\bf Evaluate Competing Designs:} differences in signal strength or time to decision are due to differences in the design. In the case that individuals were shown multiple lineups (as part of a bigger study), it is possible to correct outcome measurements for an individual's visual ability. }
\end{enumerate}

### 2012-03-29 setup.tex
typo in setup, and massaging intro
finished re-working the intro

% !TEX root = comparison.tex
\section{Experimental Setup}
We want two discuss two applications, in which we are making use of lineups to compare competing designs. 
The first one arises from a discussion of how wind direction affects efficiency of an airport. In this example, we are dealing with all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under 0.5 Million flights.

@@

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different". They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This study allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

>>>>

% !TEX root = comparison.tex
\section{Experimental Setup}~\label{experimentone}
We want two discuss two applications, in which we are making use of lineups to compare competing designs. 
The first one arises from a discussion of how wind direction affects efficiency of an airport. In this example, we are dealing with all flights \cite{rita} in and out of Seattle/Tacoma International Airport (SEA) between July 2008 and June 2011. As a measure of airport efficiency we are using time between successive wheel-events (time at wheel take-off or touch-down), which we found to be independent of airline carrier and operating hours, as long as we restricted the data to `regular' operating hours between 7 am and midnight and `regular' weather conditions -- i.e. we eliminated records associated with the top 5\% wind speed measurements \cite{noaa-weather}, leaving us with just under 0.5 Million flights.

@@

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different''. They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This study allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.
